R for Clinical Study Reports and Submission

### 2021-10-09 project-folder.Rmd
update project managment
minor fixes for ch10 ch11

To deliver A&R results of a clincial tiral project, it is a team work that typically require
collaborations from both statisticians and programmers. 
In this part, let's consider how to organize a clinical trial project as an A&R leader.

In Chapter \@ref(folder), we will discuss how to organize source code, documents, 
deliverables in an A&R clinical project. 
We recommend to use the R package folder structure.

In Chapter \@ref(manage), we will discuss a process or system development lifecycle 
to manage A&R of a clinical project. 
We recommend to follow an agile management approach to define, develop, validate and deliver work.

@@

A clearly defined clinical project folder structure can have many benefits to 
clinical development teams in an organization. 
Specifically, a well defined project folder can achieve:

- Consistency: everyone work on the same folder structure.
- Reproducibility: analysis can be executed and reproduced by different team members months/years later.  
- Automation: automatically check the integration of a project.  
- Compliance: reduce compliance issue

@@

For R users, you already benefit from a well-defined and consistent folder structure. 
That is the [R package folder structure](https://github.com/rstudio/cheatsheets/raw/master/package-development.pdf). 
Every R package developer is required to follow the same convention to organize their R
functions before the R package can be disseminated through the Comprehensive R Archive Network
(CRAN). As a user, you can easily install and use those R packages after installing from CRAN. 
There are many good resources to guide developers on R package development.
For example, [Hadley Wickmans' R packages book](https://r-pkgs.org/)

@@

## Consistency

For consistency, a well-defined folder structure with potential templates
ensures project teams organize A&R work consistently across multiple projects. 
Consistent folder structure also reduces communication cost between study team members 
and enhances the transparency of projects.

In this book, we refer to an R package as **project specific R package** 
if the purpose of an R package is to organize analysis scripts for a clinical project.

@@

Below is minimal sufficient folders and files for a project specific R package
based on the R package folder structure.

- `*.Rproj`: RStudio project file used to open RStudio project.
- `DESCRIPTION`: Metadata for a package including authors, license, dependency etc.
- `R/`: Project specific R functions.
- `vignettes/`: Analysis scripts using Rmarkdown.
- `man/`: Manual of project specific R functions.

@@

In `esubdemo` project, we saved all TLF generation scripts in previous chapters into the `vignettes/` folder.

> Here the `adam\` folder contains ADaM datasets. 
> The `tlf` folder contains output TLFs in RTF format.
> We put `adam\` and `tlf\` folder within the `vignettes` folder only for illustration purposes. 
> In a real A&R report, you may have a different location to save your input and output.

@@

The `man` folder is used to save documentation automatically generated by `roxygen2`.
A typical workflow is to add `roxygen2` documentation at the header of each function in the `R\` folder. 
Then `devtools::documents()` is used to generate all the documentation files in the `man\` folder. 
More details can be found in [Chapter 10 of the R packages book](https://r-pkgs.org/man.html).

@@

In `esubdemo` project, a reproducible environment is created when you open the 
`esubdemo.Rproj` from RStudio. 
While RStudio open `esubdemo` project, RStudio will automatically execute R code in `.Rprofile`.
So we can use `.Rprofile` to setup a reproducible environment. 
More details can be found in [https://rstats.wtf/r-startup.html](https://rstats.wtf/r-startup.html).

> `.Rprofile` is only for project specific R package. 
> A standard R package should not use `.Rprofile`.

After we open the `esubdemo` project, the code in `.Rprofile` will automatically check 
current R version is the same as we defined in `.Rprofile`

```{r, eval = FALSE}
R_version <- "4.1.1" # set up project R version
```

@@

# set up repository based on snapshot
repos     <- paste0("https://mran.microsoft.com/snapshot/", snapshot)

# Define repo URL for project specific package installation
options(repos = repos)
```

@@

In other words, all R packages installed in this RStudio project 
are based on the frozen R version at the snapshot date. 
Here it is `2021-08-06` by using MRAN server.

@@

```
Current oroject R package repository:
    https://mran.microsoft.com/snapshot/2021-08-06
```

@@

Secondly, we use `renv` to lock R package versions and saved in `renv.lock` file. 
The `renv` provides a robust and stable approach to manage R package versions 
for a project specific R packages.
An introduction of `renv` can be found on its 
[package website](https://rstudio.github.io/renv/articles/renv.html).

@@

The R code above in the `.Rprofile` initiate the `renv` running environment.  
As a user, you can use `renv::init()`, `renv::snapshot()` and `renv::restore()` 
to initialize, save and restore R packages used for the current analysis project.

@@

> `renv.lock` file and `renv/` folder are only for project specific R package. 
> A standard R package should not use `renv`.

@@

> A cloud-based R environment (e.g. RStudio Workbench) 
> can enhance the reproducibility within an organization 
> by using the same operation system, R and R package versions for an A&R project. 
> More details can be found in [https://environments.rstudio.com/](https://environments.rstudio.com/).

> A container like [docker](https://github.com/rstudio/r-docker) can further enhance 
> the reproducibility across an organization at operation system level, but beyond the scope of this book.

In conclusion, to achieve reproducibility for a project specific R package, 
a clinical project team can work under a controlled R environment 
in the same R version and R package versions defined by a repository snapshot date.

@@

We have learned a few functions in `devtools` to automatically generate content. 
Here is a short list of tools that can enhance the workflow.

- [`devtools`](https://devtools.r-lib.org/): make package development easier. 
  + A good overview can be found in [Chapter 2 of the R packages](https://r-pkgs.org/whole-game.html) book  
  + `devttols::load_all()`: load all functions in `R/` folder and running environment. 
  + `devtools::document()`: automatically create documentation using `Roxygen2`
  + `devttols::check()`: automatically perform compliance check as an R package. 
  + `devtools::build_site()`: automatically run analysis scripts in batch and create a `pkgdown` website. 
- [`usethis`](https://usethis.r-lib.org/): automates repetitive tasks that arise during project setup and development.
- [`testthat`](https://testthat.r-lib.org/): streamline testing code.
  + A discussion of using the `testthat` for an A&R project can be found in (@madhutest).  
- [`pkgdown`](https://pkgdown.r-lib.org/): generate static html documentation for an R package
  + It also allows you to run all analysis code in batch.

You may further automatically execute routines by leveraging CI/CD workflow. 
For example, the `esubdemo` project will rerun all required check and build pkgdown website
by using [github action](https://usethis.r-lib.org/reference/github_actions.html).

>>>>

To deliver A&R results of a clinical trial project, it is teamwork that typically requires
collaborations from both statisticians and programmers. 
In this part, let's consider how to organize a clinical trial project as an A&R leader.

In Chapter \@ref(folder), we will discuss how to organize source code, documents, 
deliverables in an A&R clinical project. 
We recommend using the R package folder structure.

In Chapter \@ref(manage), we will discuss a process or system development lifecycle 
to manage the A&R of a clinical project. 
We recommend following an agile management approach to define, develop, validate and deliver work.

@@

A clearly defined clinical project folder structure can have many benefits to 
clinical development teams in an organization. 
Specifically, a well-defined project structure can achieve:

- Consistency: everyone works on the same folder structure.
- Reproducibility: analysis can be executed and reproduced by different team members months/years later.  
- Automation: automatically check the integration of a project.  
- Compliance: reduce compliance issue

@@

For R users, you already benefit from a well-defined and consistent folder structure. 
That is the [R package folder structure](https://github.com/rstudio/cheatsheets/raw/master/package-development.pdf). 
Every R package developer is required to follow the same convention to organize their R
functions before the R package can be disseminated through the Comprehensive R Archive Network
(CRAN). As a user, you can easily install and use those R packages after installing from CRAN. 
There are many good resources to guide developers on R package development.
For example, Hadley Wickham's [R Packages book](https://r-pkgs.org/).

@@

## Consistency {#consistency}

For consistency, a well-defined folder structure with potential templates
ensures project teams organize A&R work consistently across multiple projects. 
Consistent folder structure also reduces communication costs between study team members 
and enhances the transparency of projects.

In this book, we refer to an R package as **project-specific R package** 
if the purpose of an R package is to organize analysis scripts for a clinical project.

@@

Below is minimal sufficient folders and files for a project-specific R package
based on the R package folder structure.

- `*.Rproj`: RStudio project file used to open RStudio project.
- `DESCRIPTION`: Metadata for a package including authors, license, dependencies, etc.
- `R/`: Project-specific R functions.
- `vignettes/`: Analysis scripts using Rmarkdown.
- `man/`: Manual of project-specific R functions.

@@

In the `esubdemo` project, we saved all TLF generation scripts in previous chapters into the `vignettes/` folder.

> Here, the `adam/` folder contains ADaM datasets. 
> The `tlf` folder contains output TLFs in RTF format.
> We put `adam/` and `tlf/` folders within the `vignettes` folder only for illustration purposes. 
> In an actual A&R report, you may have a different location to save your input and output.

@@

The `man` folder is used to save documentation automatically generated by `roxygen2`.
A typical workflow is to add `roxygen2` documentation at the header of each function in the `R/` folder. 
Then `devtools::documents()` is used to generate all the documentation files in the `man/` folder. 
More details can be found in [Chapter 10 of the R packages book](https://r-pkgs.org/man.html).

@@

In the `esubdemo` project, a reproducible environment is created when you open the 
`esubdemo.Rproj` from RStudio. 
When we open the `esubdemo` project, RStudio will execute the R code in `.Rprofile` automatically.
So we can use `.Rprofile` to set up a reproducible environment. 
More details can be found in [https://rstats.wtf/r-startup.html](https://rstats.wtf/r-startup.html).

> `.Rprofile` is only for project-specific R package. 
> A standard R package should not use `.Rprofile`.

After we open the `esubdemo` project, the code in `.Rprofile` will automatically check 
the current R version is the same as we defined in `.Rprofile`

```{r, eval = FALSE}
# set up project R version
R_version <- "4.1.1"
```

@@

# set up repository based on the snapshot date
repos <- paste0("https://mran.microsoft.com/snapshot/", snapshot)

# define repo URL for project-specific package installation
options(repos = repos)
```

@@

In other words, all R packages installed in this RStudio project 
are based on the frozen R version at the snapshot date. 
Here it is `2021-08-06` by using the MRAN server.

@@

```
Current project R package repository:
    https://mran.microsoft.com/snapshot/2021-08-06
```

@@

Secondly, we use `renv` to lock R package versions and save them in the `renv.lock` file.
`renv` provides a robust and stable approach to managing R package versions 
for project-specific R packages.
An introduction of `renv` can be found on its 
[website](https://rstudio.github.io/renv/articles/renv.html).

@@

The R code above in the `.Rprofile` initiates the `renv` running environment.  
As a user, you can use `renv::init()`, `renv::snapshot()` and `renv::restore()` 
to initialize, save and restore R packages used for the current analysis project.

@@

> `renv.lock` file and `renv/` folder are only for project-specific R package.
> A standard R package should not use `renv`.

@@

> A cloud-based R environment (e.g., RStudio Workbench)
> can enhance the reproducibility within an organization
> by using the same operating system, R version, and R package versions for an A&R project.
> More details can be found in [https://environments.rstudio.com/](https://environments.rstudio.com/).

> A container solution like [Docker](https://github.com/rstudio/r-docker) could further enhance 
> the reproducibility across an organization at the operating system level, but beyond the scope of this book.

In conclusion, to achieve reproducibility for a project-specific R package, 
a clinical project team can work under a controlled R environment 
in the same R version and R package versions defined by a repository snapshot date.

@@

We have learned a few functions in `devtools` to generate content automatically.
Here is a shortlist of tools that can enhance the workflow.

- [`devtools`](https://devtools.r-lib.org/): make package development easier. 
  + A good overview can be found in [Chapter 2 of the R packages](https://r-pkgs.org/whole-game.html) book  
  + `devttols::load_all()`: load all functions in `R/` folder and running environment. 
  + `devtools::document()`: automatically create documentation using `roxygen2`.
  + `devttols::check()`: automatically perform compliance check as an R package. 
  + `devtools::build_site()`: automatically run analysis scripts in batch and create a `pkgdown` website. 
- [`usethis`](https://usethis.r-lib.org/): automates repetitive tasks that arise during project setup and development.
- [`testthat`](https://testthat.r-lib.org/): streamline testing code.
  + A discussion of using the `testthat` for an A&R project can be found in (@madhutest).  
- [`pkgdown`](https://pkgdown.r-lib.org/): generate static HTML documentation website for an R package
  + It also allows you to run all analysis code in batch.

You may further automatically execute routines by leveraging CI/CD workflow. 
For example, the `esubdemo` project will rerun all required checks and build a pkgdown website
by using [Github Actions](https://usethis.r-lib.org/reference/github_actions.html).
